### 2020-02-25 Writing/disparity_review.tex
Edits to intro

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
$^2$IDepartment of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
$^{4}$Department of Biology, University of York, YO10 5DD, UK;
$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
$^{8}$ @@@;
$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
$^{15}$ @@@;
$^{16}$ @@@;\\
$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}

@@

\noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids or molluscs), species-rich but phenotypically conservative (e.g. rodents or nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon. Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (e.g. taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

	\item \textbf{Descriptive disparity} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace? \citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
    More recently, this approach as also been used to understand the relation between developmental processes and morphology in the evo-devo field.
    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing to successfully link the variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.

	\item \textbf{Disparity-through-time} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?

	\item \textbf{Disparity and taxonomic diversity} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling 
	\citep[e.g.][]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.

	\item \textbf{Disparity as a proxy for ecology} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
	\citep{Friedman2010-ve, Pierce2008-yr, Anderson2013-zt}.
	The links between form and function, however, are not always clear cut.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005-or}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{Brusatte2008-vx} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010-ve}.
	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005-or}.

@@

Fundamental insights into evolutionary biology have been elicited from these four types of analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?

>>>>

\noindent {\small \it 
$^1$School of Biological Sciences, University of Queensland, St. Lucia, Queensland, Australia;
$^2$Department of Life Sciences, Natural History Museum, London, Cromwell Road, London, SW7 5BD, UK;
$^{3}$School of GeoSciences, University of Edinburgh, Grant Institute, Edinburgh EH9 3FE, UK;
$^{4}$Department of Biology, University of York, YO10 5DD, UK;
$^{5}$Department of Zoology, School of Natural Sciences, Trinity College, Ireland;
$^{6}$Institut de Syst\'{e}matique, \'{E}volution, Biodiversit\'{e} (ISYEB), Muséum national d'Histoire naturelle, CNRS, Sorbonne Universit\'{e}, EPHE, Université des Antilles, 57 rue Cuvier CP39, 75005 Paris, France;
$^{7}$Ryan Institute, School of Natural Sciences, National University of Ireland Galway, Ireland;
$^{8}$ @@@;
$^{10}$School of Earth and Environment, University of Leeds, Leeds LS2 9JT, UK;
$^{11}$School of Earth Sciences, University of Bristol, Life Sciences Building, Tyndall Avenue, Bristol BS8 1TQ, UK;
$^{12}$Milner Centre for Evolution, University of Bath, BA2 7AY UK;
$^{13}$Department of Earth Sciences, University of Oxford, S Parks Road, Oxford OX1 3AN, UK;
$^{14}$School of Biological Sciences, The University of Adelaide, Adelaide, South Australia 5005, Australia.
$^{15}$ @@@;
$^{16}$ @@@;\\
$^{+}$ These authors contributed equally to the manuscript; $^{*}$ Corresponding author: guillert@tcd.ie\\}

@@

\noindent Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids or molluscs), species-rich but phenotypically conservative (e.g. rodents or nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), % NC: Can we expand Afrotheria a bit? All the other examples are using common names. Can we use common names of afrotheres? Or at least say afrotherian mammals?
or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not necessarily linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995) % NC: Seems weird only having these two refs here and no refs in the other questions. Remove?
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are comparatively more recent. 
Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} the concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (e.g. taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

@@

	\item \textbf{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace? \citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.
    More recently, this approach as also been used to understand the relationship between developmental processes and morphology in the evo-devo field.
    For example, describing disparity patterns has been successfully used to compare modules of evolution in various groups \citep{goswami2010influence,bardua2019morphological} allowing researchers to successfully link variation in shape to a group's evolutionary or developmental constraints \citep{Hipsley2017}.

	\item \textbf{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?

	\item \textbf{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. 
	Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling of the two
	\citep[e.g.][]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.

	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
	\citep{Friedman2010-ve, Pierce2008-yr, Anderson2013-zt}.
	The links between form and function, however, are not always clear cut.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005-or}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{Brusatte2008-vx} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010-ve}.
	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005-or}.

@@

Fundamental insights into evolutionary biology have been elicited from these four types of disparity analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?
